### 2014-08-27 thesis-proposal/acc.tex

% !TEX root = omar-thesis-proposal.tex
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily, 
stringstyle=\ttfamily,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS}
}
\section{Active Code Completion}\label{acc}
An abstraction might benefit from support from a variety of tools beyond the compiler. For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}. Editors for object-oriented languages provide code completion in the form of a floating menu containing  contextually-relevant variables, assignables, fields, methods, types and other code snippets. By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.

@@

In all such systems, the code completion interface has remained primarily menu-based. When an item in the menu is selected, code is inserted immediately, without further input from the developer. These systems are difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic or provide assistance beyond that which a menu can provide. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions using active types. This  makes developing and integrating a broad array of highly-specialized code generation tools directly into the editor, via the familiar code completion command, significantly simpler.% This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

@@

\begin{figure*}\label{color}
\begin{center}
\includegraphics[width=40pc]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The elaboration generated by this palette.}
\label{colorpalette}
\end{figure*}

>>>>

% !TEX root = omar-thesis-proposal.tex
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily, 
stringstyle=\ttfamily,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,toast},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS}
}
\newpage
\section{Active Code Completion}\label{acc}
\begin{figure}[h]\label{color}
\begin{center}
\includegraphics[width=40pc]{color_palette.png}\end{center}
\caption{(a) An example code completion palette associated with the \texttt{Color} class. (b) The elaboration generated by this palette.}
\label{colorpalette}
\end{figure}

An abstraction might benefit from support from a variety of tools beyond the compiler. For example, software developers today make heavy use of the code completion support found in modern source code editors  \cite{murphy_how_2006}. Code completion typically takes the form of a floating menu containing  contextually-relevant variables, assignables, fields, methods, types and other code snippets. When an item in the menu is selected, code is inserted immediately, without further input from the developer.  By navigating and selecting from this menu, developers are able to avoid many common spelling and logic errors, eliminate unnecessary keystrokes and explore unfamiliar APIs without incurring the mental overhead associated with switching to an external documentation tool.
% The way code completion menus brought documentation systems into the editor, active code completion interfaces bring external tools into the editor.

@@

These systems are difficult to extend: a fixed strategy determines the completions that are available, so library providers cannot directly specify new domain-specific or contextually-relevant logic or provide assistance beyond that which a menu can provide. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions using active types. This  makes developing and integrating a broad array of highly-specialized code generation tools directly into the editor, via the familiar code completion command, significantly simpler.% This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

@@

### 2014-08-27 thesis-proposal/background.tex

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions. For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{ScalaMacros2013}, take full control over all of the code that they enclose. This can be problematic, however, as outer macros can interfere with inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a term in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}). Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it, and to build tools that operate robustly in their presence.

@@

Another example of an active library approach to extensibility with non-local scope is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}. These languages permit libraries to extend the base syntax of the core language in a nearly arbitrary manner, and these extensions are imported transitively throughout a program. Unfortunately, this flexibility again means that  extensions are  not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same notations but back them with differing implementations. 
And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.

>>>>

More powerful and direct compile-time \emph{term rewriting mechanisms} available in some languages can also be used for optimization, as well as for introducing specialized error checking logic and implementing new abstractions. For example, typed macros, such as those in MetaML \cite{Sheard:1999:UMS}, Template Haskell \cite{SheardPeytonJones:Haskell-02} and Scala \cite{ScalaMacros2013}, take full control over all of the code that they enclose. This can be problematic, however, as outer macros can interfere with inner macros. Moreover, once a value escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope, because the output of a macro is simply a term in the underlying language (a problem fundamentally related to the problem of relying on a common intermediate language,  described in Section \ref{external-approaches}). Thus, macros represent at best a partial solution: they can be used to automate code generation, but not to globally extend the syntax or static guarantees of a language. It can also be difficult to reason about the semantics of code when any number of enclosing macros may be manipulating it (particularly if they can change its type, as in some static macro systems), and to build tools that operate robustly in their presence.

@@

Another example of an active library approach to extensibility with non-local scope is SugarJ \cite{erdweg2011sugarj} and other languages generated by Sugar* \cite{erdweg2013framework}, like SugarHaskell \cite{erdweg2012layout}. These languages permit libraries to extend the base syntax of the core language in a nearly arbitrary manner, and these extensions are imported transitively throughout a program. Unfortunately, this flexibility again means that  extensions are  not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same syntax but back them with differing implementations. 
And again, it is difficult to predict what an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.
